---
title: "Project_2_student_success"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Introduction

This project investigates how student behavioral and environmental factors relate to academic performance. We use the **Student Performance Dataset** (6,607 observations, sourced from Kaggle, updated March 2026) to answer two SMART questions:

1. Which combination of behavioral and environmental factors most strongly predicts **Final Exam Score**, and how much variance do they explain?
2. After categorizing students as passing or failing, can we accurately **identify at-risk students**, and which factors are the strongest predictors of failure?

---

# 2. Load Libraries and Data

```{r load}
# Install any missing packages before running:
# install.packages(c("ggplot2", "corrplot", "car", "pROC", "rpart", "rpart.plot", "dplyr", "caret"))

#install.packages(c("ggplot2", "corrplot", "car", "pROC", "rpart", "rpart.plot", "dplyr", "caret"))

library(ggplot2)
library(corrplot)
library(car)
library(pROC)
library(rpart)
library(rpart.plot)
library(dplyr)
library(caret)

# Load data
setwd("~/Desktop/DS_Project_2")
list.files()
df <- read.csv("student_dataset (1).csv")

# Quick look
dim(df)
head(df)
str(df)
```

---

# 3. Data Cleaning & Preparation

```{r clean}
# Check for missing values
colSums(is.na(df))

# Convert categorical variables to factors
df$Parental_Involvement      <- as.factor(df$Parental_Involvement)
df$Access_to_Resources       <- as.factor(df$Access_to_Resources)
df$Extracurricular_Activities <- as.factor(df$Extracurricular_Activities)
df$Motivation_Level          <- as.factor(df$Motivation_Level)
df$Internet_Access           <- as.factor(df$Internet_Access)

# Confirm structure
str(df)
```

---

# 4. Exploratory Data Analysis (EDA)

## 4.1 Summary Statistics

```{r summary}
summary(df)
```

## 4.2 Distribution of Final Exam Score

```{r hist}
ggplot(df, aes(x = Final_Exam_Score)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Final Exam Score",
       x = "Final Exam Score", y = "Count") +
  theme_minimal()
```

## 4.3 Correlation Matrix (Numeric Variables)

```{r corrplot}
# Select only numeric columns
num_vars <- df %>% select(Hours_Studied, Attendance, Sleep_Hours,
                           Previous_Scores, Tutoring_Sessions, Final_Exam_Score)

cor_matrix <- cor(num_vars)
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix", mar = c(0,0,1,0))
```

## 4.4 Scatterplots: Key Predictors vs Final Exam Score

```{r scatter}
# Hours Studied vs Final Score
ggplot(df, aes(x = Hours_Studied, y = Final_Exam_Score)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Hours Studied vs Final Exam Score") +
  theme_minimal()

# Previous Scores vs Final Score
ggplot(df, aes(x = Previous_Scores, y = Final_Exam_Score)) +
  geom_point(alpha = 0.3, color = "darkorange") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Previous Scores vs Final Exam Score") +
  theme_minimal()

# Attendance vs Final Score
ggplot(df, aes(x = Attendance, y = Final_Exam_Score)) +
  geom_point(alpha = 0.3, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Attendance vs Final Exam Score") +
  theme_minimal()
```

## 4.5 Boxplots: Categorical Variables vs Final Exam Score

```{r boxplots}
# Parental Involvement
ggplot(df, aes(x = Parental_Involvement, y = Final_Exam_Score, fill = Parental_Involvement)) +
  geom_boxplot() +
  labs(title = "Final Exam Score by Parental Involvement") +
  theme_minimal() + theme(legend.position = "none")

# Access to Resources
ggplot(df, aes(x = Access_to_Resources, y = Final_Exam_Score, fill = Access_to_Resources)) +
  geom_boxplot() +
  labs(title = "Final Exam Score by Access to Resources") +
  theme_minimal() + theme(legend.position = "none")

# Motivation Level
ggplot(df, aes(x = Motivation_Level, y = Final_Exam_Score, fill = Motivation_Level)) +
  geom_boxplot() +
  labs(title = "Final Exam Score by Motivation Level") +
  theme_minimal() + theme(legend.position = "none")

# Internet Access
ggplot(df, aes(x = Internet_Access, y = Final_Exam_Score, fill = Internet_Access)) +
  geom_boxplot() +
  labs(title = "Final Exam Score by Internet Access") +
  theme_minimal() + theme(legend.position = "none")

# Extracurricular Activities
ggplot(df, aes(x = Extracurricular_Activities, y = Final_Exam_Score, fill = Extracurricular_Activities)) +
  geom_boxplot() +
  labs(title = "Final Exam Score by Extracurricular Activities") +
  theme_minimal() + theme(legend.position = "none")
```

---

# 5. SMART Question 1: Multiple Linear Regression

**Which combination of factors most strongly predicts Final Exam Score?**

## 5.1 Full Model

```{r lm_full}
lm_full <- lm(Final_Exam_Score ~ ., data = df)
summary(lm_full)
```

## 5.2 Variable Selection (Stepwise AIC)

```{r stepwise}
lm_step <- step(lm_full, direction = "both", trace = 0)
summary(lm_step)
```

## 5.3 Model Diagnostics

```{r diagnostics}
par(mfrow = c(2, 2))
plot(lm_step)
par(mfrow = c(1, 1))
```

```{r vif}
# Check for multicollinearity (VIF)
# Only works cleanly if no aliased coefficients remain
vif(lm_step)
```

## 5.4 Interpretation

```{r lm_results}
# Coefficients and confidence intervals
coef(summary(lm_step))
confint(lm_step)

# Adjusted R-squared
summary(lm_step)$adj.r.squared
```

> **Interpretation:** The adjusted R² tells us what proportion of variance in Final Exam Score is explained by the model. Significant predictors (p < 0.05) are the key behavioral/environmental factors. Positive coefficients mean that variable increases predicted score; negative means the opposite.

---

# 6. SMART Question 2: Logistic Regression & Decision Tree

**Can we identify at-risk (failing) students, and which factors predict failure?**

## 6.1 Create Pass/Fail Variable

```{r passfail}
# Define passing as Final Exam Score >= 60
df$Pass <- ifelse(df$Final_Exam_Score >= 60, 1, 0)
df$Pass <- as.factor(df$Pass)

table(df$Pass)
prop.table(table(df$Pass))
```

## 6.2 Train/Test Split

```{r split}
set.seed(123)
train_index <- createDataPartition(df$Pass, p = 0.75, list = FALSE)
train <- df[train_index, ]
test  <- df[-train_index, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test), "\n")
```

## 6.3 Logistic Regression

```{r logistic}
# Drop Final_Exam_Score from predictors (we created Pass from it)
log_model <- glm(Pass ~ Hours_Studied + Attendance + Sleep_Hours +
                   Previous_Scores + Tutoring_Sessions +
                   Parental_Involvement + Access_to_Resources +
                   Motivation_Level + Internet_Access +
                   Extracurricular_Activities,
                 data = train, family = binomial)

summary(log_model)
```

## 6.4 Stepwise Variable Selection for Logistic Model

```{r log_step}
log_step <- step(log_model, direction = "both", trace = 0)
summary(log_step)
```

## 6.5 Predictions & Confusion Matrix

```{r confusion}
# Predict on test set
log_probs <- predict(log_step, newdata = test, type = "response")
log_preds <- ifelse(log_probs > 0.5, 1, 0)
log_preds <- as.factor(log_preds)

# Confusion Matrix
confusionMatrix(log_preds, test$Pass, positive = "1")
```

## 6.6 ROC Curve & AUC

```{r roc}
roc_obj <- roc(as.numeric(test$Pass), log_probs)
plot(roc_obj, col = "steelblue", main = "ROC Curve - Logistic Regression")
auc(roc_obj)
```

## 6.7 Decision Tree

```{r tree}
tree_model <- rpart(Pass ~ Hours_Studied + Attendance + Sleep_Hours +
                      Previous_Scores + Tutoring_Sessions +
                      Parental_Involvement + Access_to_Resources +
                      Motivation_Level + Internet_Access +
                      Extracurricular_Activities,
                    data = train, method = "class",
                    control = rpart.control(cp = 0.01))

# Plot the tree
rpart.plot(tree_model, type = 4, extra = 104,
           main = "Decision Tree: Predicting Pass/Fail")
```

## 6.8 Decision Tree Predictions & Evaluation

```{r tree_eval}
tree_preds <- predict(tree_model, newdata = test, type = "class")

confusionMatrix(tree_preds, test$Pass, positive = "1")
```

## 6.9 Variable Importance (Decision Tree)

```{r varimp}
importance <- tree_model$variable.importance
importance_df <- data.frame(Variable = names(importance), Importance = importance)
importance_df <- importance_df[order(-importance_df$Importance), ]

ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance - Decision Tree",
       x = "Variable", y = "Importance") +
  theme_minimal()
```

---

# 7. Model Comparison

```{r compare}
# Logistic Regression accuracy
log_acc <- mean(log_preds == test$Pass)

# Decision Tree accuracy
tree_acc <- mean(tree_preds == test$Pass)

comparison <- data.frame(
  Model    = c("Logistic Regression", "Decision Tree"),
  Accuracy = round(c(log_acc, tree_acc), 4)
)

print(comparison)
```


